# Alpha diversity demo

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Alpha diversity estimation


First let`s load the required packages and data set


```{r load}
library(mia)
library(miaViz)
library(tidyverse)
# library(vegan)

tse <- read_rds("data/Tengeler2020/tse.rds")

tse

```
Then let's estimate multiple diversity indices.

```{r estimate diversity}

?estimateDiversity

tse <- estimateDiversity(tse, 
                              index = c("shannon","gini_simpson","faith"),
                              name = c("shannon","gini_simpson","faith"))
head(colData(tse))

```

We can see that the variables are included in the data.
Similarly, let's calculate richness indices.

```{r estimate richness}

tse <- estimateRichness(tse, 
                              index = c("chao1","observed"))
head(colData(tse))

```

## Visualizing alpha diversity 

We can plot the distributions of individual indices:

```{r distributions}
#individual plot
p <- as_tibble(colData(tse)) %>% 
  ggplot(aes(shannon)) +
  geom_histogram() 

print(p)

#multiple plots

p <- as_tibble(colData(tse)) %>% 
  pivot_longer(cols = c("shannon","gini_simpson","faith","chao1","observed"), names_to = "index", values_to = "alpha") %>% 
  ggplot(aes(alpha)) +
  geom_histogram() +
  facet_wrap(vars(index), scales = "free")


print(p)


```

and the correlation between indices:

```{r scatterlots, fig.width=13,fig.height=12}

p <- as_tibble(colData(tse)) %>% 
  pivot_longer(cols = c("shannon","gini_simpson","faith","chao1","observed"), names_to = "index", values_to = "alpha") %>% 
  full_join(.,., by = "sample_name") %>% 
  ggplot( aes(x = alpha.x, y = alpha.y)) + 
  geom_point() +
  geom_smooth() +
  facet_wrap(index.x ~ index.y, scales = "free")

print(p)

```

## Comparing alpha diversity 

It is often interesting to look for any group differences:


```{r boxplots}

p <- as_tibble(colData(tse)) %>% 
  pivot_longer(cols = c("shannon","gini_simpson","faith","chao1","observed"), names_to = "index", values_to = "alpha") %>% 
  ggplot( aes(x = patient_status, y = alpha)) + 
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(alpha =0.5) +
  facet_wrap(vars(index), scales = "free")

print(p)

```

Moreover, we can test the group differences by parametric or non-parametric tests:

```{r comparison}

df1 <- as_tibble(colData(tse)) %>% 
  pivot_longer(cols = c("faith","chao1","observed"), names_to = "index", values_to = "alpha") %>% 
  group_by(index) %>% 
  nest() %>% 
  mutate(test_pval = map_dbl(data, ~ t.test(alpha ~ patient_status, data = .x)$p.value)) %>% 
  mutate(test = "ttest" ) 

df2 <- as_tibble(colData(tse)) %>% 
  pivot_longer(cols = c("shannon","gini_simpson"), names_to = "index", values_to = "alpha") %>% 
  group_by(index) %>% 
  nest() %>% 
  mutate(test_pval = map_dbl(data, ~ wilcox.test(alpha ~ patient_status, data = .x)$p.value))%>% 
  mutate(test = "wilcoxon" ) 

df <- rbind(df1,df2) %>% select(-data) %>% arrange(test_pval) %>% ungroup()

df

```
End of the demo.


## Exercises

Do "Alpha diversity basics" from the [exercises](https://microbiome.github.io/OMA/exercises.html).
