# Beta diversity demo 

```{r setup, include=FALSE, message= F}
library(mia)
library(miaViz)
library(tidyverse)
library(scater)

#### loading data 
tse <- read_rds("data/Tengeler2020/tse.rds")
```

## Visualizations

Lets generate ordination plots with different methods and transformations.


```{r viz1}

#### calculating Bray Curtis dissimilarity and PCoA

tse <- transformSamples(tse, method = "relabundance")
tse <- runMDS(tse, FUN = vegan::vegdist, method = "bray", name = "PCoA_BC", exprs_values = "relabundance")


p <- plotReducedDim(tse, "PCoA_BC", colour_by = "patient_status")

# Add explained variance for each axis
e <- attr(reducedDim(tse, "PCoA_BC"), "eig");
rel_eig <- e/sum(e[e>0])          
p <- p + labs(x = paste("PCoA 1 (", round(100 * rel_eig[[1]],1), "%", ")", sep = ""),
              y = paste("PCoA 2 (", round(100 * rel_eig[[2]],1), "%", ")", sep = ""))

print(p)

#### Aitchinson distances and PCA

tse <- transformSamples(tse, method = "clr", pseudocount = 1)

tse <- runMDS(tse, FUN = vegan::vegdist, name = "MDS_euclidean",
              method = "euclidean", exprs_values = "clr")

p <- plotReducedDim(tse, "MDS_euclidean", colour_by = "patient_status_vs_cohort")

# Add explained variance for each axis
e <- attr(reducedDim(tse, "MDS_euclidean"), "eig");
rel_eig <- e/sum(e[e>0])          
p <- p + labs(x = paste("Axis 1 (", round(100 * rel_eig[[1]],1), "%", ")", sep = ""),
              y = paste("Axis 2 (", round(100 * rel_eig[[2]],1), "%", ")", sep = ""))

print(p)
```


PCA is a subtype of MDS with Euclidean distances, below is a different alternative for running the same analysis.


```{r viz2}

# alternative method 

tse <- runPCA(tse, name = "PCA", exprs_values = "clr", ncomponents = 10)
plotReducedDim(tse, "PCA", colour_by = "patient_status_vs_cohort")

```


One can use also ggplot for ordination plots for the flexible adaptablity.


```{r viz ggplot}


dis <- vegan::vegdist(t(assays(tse)$counts), method = "jaccard")

# principal coordinate analysis
jaccard_pcoa <- ecodist::pco(dis)

# a data frame from principal coordinates and groupng variable
jaccard_pcoa_df <- data.frame(pcoa1 = jaccard_pcoa$vectors[,1], 
                                pcoa2 = jaccard_pcoa$vectors[,2],
                              patient_status_vs_cohort =  colData(tse)$patient_status_vs_cohort)


# plot
jaccard_plot <- ggplot(data = jaccard_pcoa_df, aes(x=pcoa1, y=pcoa2, color = patient_status_vs_cohort)) +
  geom_point() +
  labs(x = paste("Axis 1 (", round(100 * jaccard_pcoa$values[[1]] / sum(jaccard_pcoa$values), 1), "%", ")", sep = ""),
       y = paste("Axis 2 (", round(100 * jaccard_pcoa$values[[2]] / sum(jaccard_pcoa$values), 1), "%", ")", sep = ""),
       title = "Jaccard PCoA") +
  theme(title = element_text(size = 12)) +
  theme_light()

jaccard_plot

```

## Hypothesis testing

PERMANOVA with the function adonis is most commonly used to detect differences in multivariate data. 
adonis function was recently updated with slightly different functionality. Now the adonis2 allows independent analysis of terms.

```{r permanova}


variable_names <- c("patient_status", "cohort")

tse_genus <- agglomerateByRank(tse, "Genus")

# Apply relative transform
tse_genus <- transformSamples(tse_genus, method = "relabundance")


set.seed(12346)
# We choose 99 random permutations for speed. Consider applying more (999 or 9999) 

assay <- t(assay(tse_genus,"relabundance"))

mod <- paste("assay ~", paste(variable_names, collapse="+")) %>% as.formula()

permanova2 <- vegan::adonis2(mod,
                     by = "margin", # each term analyzed individually
                     data = colData(tse),
                     method = "bray",
                     permutations = 99)

print(permanova2)

# older adonis for reference
permanova <- vegan::adonis(mod,
                            #by = "margin", # each term analyzed sequentially
                            data = colData(tse),
                            method = "bray",
                            permutations = 99)

permanova$aov.tab


```


With older adonis version one cam calculate top coefficients driving the differences between groups.


```{r permanova coef}


# older adonis supplies the coefficients
coef <- coefficients(permanova)["cohort1",]
top.coef <- sort(head(coef[rev(order(abs(coef)))],20))

# plot 
top_taxa_coeffient_plot <- ggplot(data.frame(x = top.coef,
                                             y = factor(names(top.coef),
                                                        unique(names(top.coef)))),
                                  aes(x = x, y = y)) +
  geom_bar(stat="identity") +
  labs(x="", y="", title="Top Taxa") +
  theme_bw()

top_taxa_coeffient_plot

```

### Testing the differences in dispersion

PEMRANOVA doesn't differentiate between different within-group variation, i.e. dispersion, or the mean differences between groups, i.e. the location of the centroid. 
Follow-up testing can be done with PERMDISP2 implemented in the vegan package.


```{r dispersion}

dis <- vegan::vegdist(t(assays(tse)$counts), method = "bray")
b <- vegan::betadisper(dis, colData(tse)$cohort)
print(anova(b))


# boxplor for distances to centroid
p <- cbind(distance = as.numeric(b$distances),
          cohort = colData(tse)$cohort) %>% 
  as_tibble() %>% 
  mutate(distance = as.numeric(distance)) %>% 
  ggplot(aes(cohort, distance)) + 
  geom_boxplot() +
  theme_light()

print(p)

```

End of the demo.



## Exercises

Do "Beta diversity" from the [exercises](https://microbiome.github.io/OMA/exercises.html).